---
title: "HW3: Predicting IMDb Ratings for *The Office* Episodes"
output:
  html_document:
    toc: true
---


### Instructions

1. All assignments are individual!
2. This assignment should be uploaded as one html file (after knit) before the next Tutorial.
3. Questions about the assignment should be asked using the Assignments forums. 
5. Write your name and id at the top of the file 


**Exercise overview:** Build, document, and evaluate a regression model that predicts the IMDb rating of each episode of *The Office* using features you engineer from the raw dialogue and episode metadata. You are free to choose **any modelling approach** and **any feature set** you believe will perform well.

  
**Dataset.** `theoffice` from the **schrute** package.

---

## 0 Setup

```{r setup, include = FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE, echo = TRUE)
set.seed(123)   # make your work reproducible!
```

Install/load the packages you will need: **tidyverse**, **lubridate**, **tidymodels**, **GGally**, and any others you choose (e.g., **vip** for variable‑importance plots).

```{r packages}
# TODO: load libraries here
library(tidyverse)
library(lubridate)
library(tidymodels)
library(GGally)
library(vip)
```

---

## 1 Data preparation 

Construct a data frame `office_df` with the following columns:

| variable | description |
|----------|-------------|
| `imdb_rating`  | numeric rating at IMDb (target) |
| `total_votes`  | number of IMDb votes (will be log‑transformed) |
| `season` & `episode` | numeric identifiers |
| `halloween`, `valentine`, `christmas` | factor indicators (1/0) for holiday episodes |
| `michael_ep`   | 1 if Michael Scott appears (season ≤ 7), else 0 |
| `lines_jim`, `lines_pam`, `lines_michael`, `lines_dwight` | proportion of lines spoken by each character |



```{r data-prep}
# TODO: build office_df here
library(schrute)
data("theoffice")

#Count lines for main characters per episode
lines_per_character <- theoffice %>%
  filter(character %in% c("Jim", "Pam", "Michael", "Dwight")) %>%
  group_by(season, episode, character) %>%
  summarise(lines = n(), .groups = "drop")

#Calculate total lines per episode
total_lines_per_episode <- lines_per_character %>%
  group_by(season, episode) %>%
  summarise(total_lines = sum(lines), .groups = "drop")

#Calculate proportion of lines per character
lines_proportion <- lines_per_character %>%
  left_join(total_lines_per_episode, by = c("season", "episode")) %>%
  mutate(prop = lines / total_lines) %>%
  select(season, episode, character, prop)

# Pivot wider and rename columns
lines_wide <- lines_proportion %>%
  pivot_wider(
    names_from = character,
    values_from = prop,
    values_fill = 0
  ) %>%
  rename(
    lines_jim = Jim,
    lines_pam = Pam,
    lines_michael = Michael,
    lines_dwight = Dwight
  )

#IMDb ratings and votes
ratings_df <- theoffice %>%
  distinct(season, episode, imdb_rating, total_votes)

# Identify holiday episodes
holiday_df <- theoffice %>%
  distinct(season, episode, episode_name) %>%
  mutate(
    halloween = ifelse(grepl("halloween", episode_name, ignore.case = TRUE), 1, 0),
    christmas = ifelse(grepl("christmas", episode_name, ignore.case = TRUE), 1, 0),
    valentine = ifelse(grepl("valentine", episode_name, ignore.case = TRUE), 1, 0)
  ) %>%
  select(-episode_name)

#Create Michael Scott appearance indicator
michael_df <- theoffice %>%
  distinct(season, episode) %>%
  mutate(michael_ep = ifelse(season <= 7, 1, 0))

#Merge all into office_df
office_df <- lines_wide %>%
  left_join(ratings_df, by = c("season", "episode")) %>%
  left_join(holiday_df, by = c("season", "episode")) %>%
  left_join(michael_df, by = c("season", "episode"))

glimpse(office_df)
```

---

## 2 Train / test split 

* Create an 80/20 stratified split on `imdb_rating`.

```{r split}
set.seed(123)
office_split <- initial_split(office_df,prop = 0.80)
train_data <- training(office_split)
test_data <- testing(office_split)
glimpse(train_data)
glimpse(test_data)
# TODO: office_split, office_train, office_test
```

---

## 3 Recipe 

* **Label:** `imdb_rating` (numeric).
* **Predictors:** use **any features you create in Section 1**—feel free to start with the full set or experiment with subsets/interactions.  
* **Pre‑processing:** add any steps you find helpful. Does the dataset include imbalanced classes? If so, how would you address that?
In this recipe, I prepared the data to be ready for modeling. Here are the steps I took:

Removed holiday indicators (halloween, christmas, and valentine) because they were extremely unbalanced — only a few episodes had them, so I didn’t want them to add noise to the model.

Applied a log transformation to the total_votes column. This helps reduce skewness since the number of votes varies a lot between episodes.

Converted categorical variables to dummy variables using step_dummy(), which is useful for modeling since many algorithms don’t work directly with text or factors.

```{r recipe}
# TODO: rating_rec <- 
office_rec <- recipe(
  imdb_rating ~ .,      
  data = train_data   
)

ggplot(train_data, aes(imdb_rating)) +
  geom_histogram(binwidth = 0.2, fill = "steelblue", color = "white") +
  labs(title = "Distribution of IMDb Ratings")
table(office_df$valentine)
table(office_df$halloween)
table(office_df$christmas)
table(office_df$season)
summary(office_df$lines_dwight)

office_rec_2 <- recipe(imdb_rating ~ ., data = train_data) %>%
  step_rm(halloween, christmas, valentine) %>% 
  step_log(total_votes, base = 10) %>%        
  step_dummy(all_nominal_predictors()) %>%      
  step_zv(all_predictors()) 
office_prep <- prep(office_rec_2)
bake(office_prep, new_data = NULL) %>% colnames()


```

---

## 4 Model specification & workflow 

* Briefly explain *why* its a regression problem and not a classification 
imdb_rating is a numeric continuous variable, so regression is appropriate.
* Create a model specification.  
* Combine the recipe and model into a `workflow`.

```{r workflow}
office_mod <- linear_reg() %>%
  set_engine("lm")
office_wf <- workflow() %>%
  add_model(office_mod) %>%
  add_recipe(office_rec_2)
# TODO
```


---

## 5 Fit model & interpret 

* Fit the workflow on the **training** data.  
* Evaluate variable importance.  
* Provide a brief interpretation of at least **two** influential predictors.

We trained a linear regression model on the training data. The most important predictors were total_votes (log-transformed), which had a strong positive effect on rating, and michael_ep, showing that episodes with Michael Scott generally got higher ratings. Other variables like season and lines_dwight also had some effect. lines_pam was dropped due to linear dependence, since the character proportions sum to 1 — the model can infer it from the others.

```{r fit}
office_fit <- office_wf  %>%
fit(data = train_data)
nrow(tidy(office_fit))
tidy(office_fit)%>% print(n = 9)

train_data %>%
  select(lines_jim, lines_pam, lines_michael, lines_dwight) %>%
  cor()

# TODO
```


---

## 6 Evaluate on the test set 

* Predict on `office_test` and evaluate the results using fit evaluation metrics.

```{r evaluate}
# TODO
office_preds <- predict(office_fit, test_data) %>%
  bind_cols(test_data)
metrics(office_preds, truth = imdb_rating, estimate = .pred)

```

Discuss whether the model meets your expectations.

---

## 7 Reflection 

Yes, the model mostly met my expectations. I had a feeling it would perform reasonably well, especially since we saw a strong relationship between the number of votes and the IMDb rating. That, combined with other features like the proportion of lines by key characters, gave the model a good foundation. The RMSE and R² confirm that it's picking up on meaningful patterns, even with a relatively simple setup. If I had more time, I would also try running the model with different seeds and averaging the results to improve robustness in addition I would explore richer feature engineering—especially from the actual dialogue. For example, I’d try using a language model (LLM) to extract sentiment or key themes from highly rated episodes. That could give the model deeper insight into what makes an episode resonate with viewers.

Describe **one** concrete improvement you would attempt if you had more time.

---

